---
title: "Bakin Fig.1"
author: "Mohammed Aruwaili"
date: "11/13/2020"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
editor_options:
  chunk_output_type: console
---

```{r, warning = F, message = F, echo = F}
knitr::opts_chunk$set(echo = FALSE, fig.width=16,fig.height=4)

# options
#options(stringsAsFactors = F)
#knit_hooks$set(optipng = hook_optipng)
#knit_hooks$set(pngquant = hook_pngquant)
#knitr::opts_chunk$set(message = F, warning = F, fig.align = "center", echo = F)



labelit<-function(text,cex=NULL,...)
{
    ds <- dev.size("in")
    # xy coordinates of device corners in user coordinates
    x <- grconvertX(c(0, ds[1]), from="in", to="user")
    y <- grconvertY(c(0, ds[2]), from="in", to="user")  
  
      fig <- par("fig")
      dx <- (x[2] - x[1])
      dy <- (y[2] - y[1])
      x <- x[1] + dx * fig[1:2]
      y <- y[1] + dy * fig[3:4]
      
      sw <- strwidth(text, cex=cex) * 60/100
      sh <- strheight(text, cex=cex) * 60/100
      
      x1<-x[1] + sw
      y1<-y[2] - sh
      
      text(x1, y1, text, cex=cex, xpd=NA, ...)
  
}

library(ComplexHeatmap)
library(circlize)
library(knitr)
library(ggplot2)
library(plyr)
```


```{r}
col<-c("dodgerblue3", "firebrick3","tan1","limegreen")
palette(col)


setwd("~/TCGA/Breast Cancer")
META<-t(read.csv("2020-0624-MAdata/Fig.E/final list of RRG, BER, MMR genes.csv",nrow=3,header=T,row.names=1))
META<-as.data.frame(META)
META$MUT<-as.numeric(as.character(META$Mutation_Count))
META$PAM<-META$PAM50_subtype
META$TP53<-as.character(META$TP53)
META<-META[META$PAM %in% c("Basal-like", "HER2-enriched", "Luminal A", "Luminal B"),]
#levels(META$PAM)[5]<-NA
#levels(META$PAM)[levels(META$PAM)%in%c("Normal-like","")]<-NA


# change the P53 status to Wild Type amd Mutation
META[META=="MUT;"]<-"Mutation"
META[META=="HOMDEL;"]<-"Wild Type"
META[META=="MUT; HOMDEL;"]<-"Mutation"
META[META=="HOMDEL; MUT;"]<-"Mutation"
META[META=="AMP;"]<-"Wild Type"
META[META==""] <- "Wild Type"

#chang2 HER2-emriched to HER2
META$PAM<-as.character(META$PAM)
META[META=="HER2-enriched"]<- "HER2"

MUTS<-t(read.csv("2020-0624-MAdata/Fig.E/final list of RRG, BER, MMR genes.csv",row.names=1,nrow=29))
MUTS<-as.data.frame(MUTS[rownames(META),])

S<-read.csv("2020-0624-MAdata/Fig.A/S_phase.csv",skip=3,header=F,row.names=1)
colnames(S)<-colnames(read.csv("2020-0624-MAdata/Fig.A/S_phase.csv",nrows=1,header=T,row.names=1))

M<-read.csv("2020-0624-MAdata/Fig.B/M_phases.csv",skip=3,
            header=F,row.names=1)
colnames(M)<-colnames(read.csv("2020-0624-MAdata/Fig.B/M_phases.csv",nrows=1,header=T,row.names=1))

BER<-read.csv("2020-0624-MAdata/Fig.C/BER genes.csv",
              skip=3, header=F,row.names=1)
colnames(BER)<-colnames(read.csv("2020-0624-MAdata/Fig.C/BER genes.csv",nrows=1,header=T,row.names=1))

MMR<-read.csv("2020-0624-MAdata/Fig.D/MMR genes.csv",
              skip=3,header=F,row.names=1)
colnames(MMR)<-colnames(read.csv("2020-0624-MAdata/Fig.D/MMR genes.csv",nrows=1,header=T,row.names=1))

RRG<-read.csv("2020-0624-MAdata/Fig.F/RRG.csv",
              skip=2,header=F,row.names=1)
colnames(RRG)<-colnames(read.csv("2020-0624-MAdata/Fig.F/RRG.csv",nrows=1,header=T,row.names=1))



SS<-apply(S,2,mean,na.rm=TRUE)[rownames(META)]
MM<-apply(M,2,mean,na.rm=TRUE)[rownames(META)]
BR<-apply(BER,2,mean,na.rm=TRUE)[rownames(META)]
MR<-apply(MMR,2,mean,na.rm=TRUE)[rownames(META)]
RG<-apply(RRG,2,mean,na.rm=TRUE)[rownames(META)]


# PAM Subtypes colors
col1 = ifelse(META$PAM=="Basal-like", "dodgerblue3",
            ifelse(META$PAM=="HER2", "firebrick3",
                   ifelse(META$PAM== "Luminal A", "tan1","limegreen")))

col2<-rgb(c(0,255,255,0),
          c(0,0,165,255),
          c(255,0,0,0),
          max=255,
          alpha=128)

col4= c("dodgerblue3", "firebrick3","tan1","limegreen")
#PAM Classification. 
PAM1= c("Basal-like", "HER2", "Luminal A", "Luminal B")

#P53 Classification
META$TP53 <- factor(META$TP53, levels = c("Wild Type", "Mutation"))



```



A z-score for a sample indicates the number of standard deviations away from the mean of expression in the reference.  

The formula is : 
z = (expression in tumor sample - mean expression in reference sample) / standard deviation of expression in reference sample

"For mRNA expression data, They typically compute the relative expression of an individual gene and tumor to the gene's expression distribution in a reference population. That reference population is either all tumors that are diploid for the gene in question, or, when available, normal adjacent tissue."


```{r}
# create the plot margin
#bitmap("S, M , TMB gene expression.png", width = 8, height = 4.5, res = 300)

layout(matrix(c(1,4,2,5,3,6),2,3))
par(mar=c(5,6,2,1))


#Panel A: S phase pathway boxplot based on molecular PAM Subtype. 
boxplot(SS~META$PAM,col=col2,border=col2,xlab="PAM50 Molecular Subtypes", ylab="S Gene Expression", cex.axis =1.9, cex.lab=2, cex.main=1, ylim=c(-3,4))
#labelit("A",cex=1.5, font=2)

#Panel B: M phase pathway boxplot based on molecular PAM Subtype. 
boxplot(MM~META$PAM,col=col2,border=col2,xlab="PAM50 Molecular Subtypes", ylab="M Gene Expression", cex.axis =1.9, cex.lab=2, cex.main=1, ylim=c(-3,4))
#labelit("B",cex=1.5, font=2)

#Panel C: Tumor Mutational Borden based on molecular PAM Subtype.
boxplot(META$MUT~META$PAM,boxfill=col2,border=col2,xlab="PAM50 Molecular Subtypes", ylab="Tumor Mutational Burden", cex.axis =1.9, cex.lab=2, cex.main=1.5, ylim= c(0,300))
#labelit("C",cex=1.5, font=2)



META$TP53 <- factor(META$TP53, levels = c("Wild Type", "Mutation"))
col3 = ifelse(META$TP53=="Mutation","firebrick3", "dodgerblue3")

# Panel D:boxplot of SS of TNBC based on TP53 Status
boxplot(SS~META$TP53,col=col3,xlab="P53 Status", ylab="S Gene Expression",cex.axis =1.9, cex.lab=2, cex.main=1.5, ylim=c(-3,4.5))
#labelit("D",cex=2, font=2)
t1<- t.test(SS~I(META$TP53=="Wild Type")) #p<0.001
t1
# for pvals
axis(3,at=1:2,line=-1.3,lab=c("",""),tcl=1)
text(x=1.5,y=0,label="p=2.2e-16",srt=4,adj=30,cex=3)

# Panel E:boxplot of MM of TNBC based on TP53 Status
boxplot(MM~META$TP53,col=col3,xlab="P53 Status", ylab="M Gene Expression",cex.axis =1.9, cex.lab=2, cex.main=1.5, ylim=c(-3,4.5))
#labelit("E",cex=2, font=2)
t2 <- t.test(MM~I(META$TP53=="Wild Type")) #p<0.001
t2
# for pvals
axis(3,at=1:2,line=-1.3,lab=c("",""),tcl=1)
text(x=1.5,y=444,label="p=2.2e-16",srt=4,adj=30,cex=3)

# Panel F:boxplot of TMB of TNBC based on TP53 Status
boxplot(META$MUT~META$TP53,col=col3,xlab="P53 Status", ylab="Tumor Mutational Burden",cex.axis =1.9, cex.lab=2, cex.main=1.5, ylim= c(0,350))
#labelit("F",cex=2, font=2)
t3 <- t.test(META$MUT~I(META$TP53=="Wild Type")) #p<0.001
t3
# for pvals
axis(3,at=1:2,line=-1.3,lab=c("",""),tcl=1)
text(x=1.5,y=444,label="p=2.2e-16",srt=4,adj=30,cex=3)

#dev.off()
#include_graphic("S, M , TMB gene expression.png", dpi=300)

```


```{r}

# Correlation Between BER, MMR and RRG Gene Sets
layout(matrix(c(1,2,3),1,3))
par(mar=c(5,6,2,1))
setx<-c("POLD3","SSBP1","EXO1","RFC4","PCNA","RFC2","MSH6","MSH2","POLD1")
set2<-c("MBD4","POLD1","LIG1","PCNA","NEIL3","FEN1","UNG","TDG")
sety<-c("BUB1","BUB3","CDC25B","CDC25A","BIRC5","CCNE2","CCNB1","CCNB2","CCNF","CDC20","CDC6","CDCA3","CENPE","CENPF","CKS1B","DTL","MCM2","NCN6","PLK1")

BR<-apply(BER[set2,],2,mean,na.rm=TRUE)[rownames(META)]
MR<-apply(MMR[setx,],2,mean,na.rm=TRUE)[rownames(META)]
RG<-apply(RRG[sety,],2,mean,na.rm=TRUE)[rownames(META)]

#Panel D :correlation plot of BER/MMR expression
plot(BR,MR,col=col1,pch=19,xlab="BER Genes Expression",ylab="MMR Genes Expression", main="BER/MMR Gene Pathway", cex.axis =1.9, cex.lab=1.8, cex.main=1.5)
legend("topleft", legend=PAM1,col=col2,pch=19, bty="n",y.intersp=0.8,x.intersp=0.8,text.width=0.1, cex=1.2)
#labelit("D",cex=1.5, font=2)

#Panel E:correlation plot of BER/MMR expression
plot(MR,RG,col=col1,pch=19,xlab="MMR Genes Expression",ylab="RRG Genes Expression", main="MMR/RRGGene Pathway", cex.axis =1.9, cex.lab=1.8, cex.main=1.5)
legend("topleft", legend=PAM1,col=col2,pch=19, bty="n",y.intersp=0.8,x.intersp=0.8,text.width=0.1, cex=1.2)
#labelit("E",cex=1.5, font=2)

#Panel F:correlation plot of BER/MMR expression
plot(BR,RG,col=col1,pch=19,xlab="BER Genes Expression",ylab="RRG Genes Expression", main="BER/RRG Gene Pathway", cex.axis =1.9, cex.lab=1.8, cex.main=1.5)
legend("topleft", legend=PAM1,col=col2,pch=19, bty="n",y.intersp=0.8,x.intersp=0.8,text.width=0.1, cex=1.2)
#labelit("F",cex=1.5, font=2)

```


```{r}

layout(matrix(c(1,4,2,5,3,6),2,3))
par(mar=c(5,6,2,1))

# Panel A: MMR and BER pathway scores based on TP53
plot(BR,MR,pch=19,col=col3, type = "p", xlab="BER Gene Expression",ylab="MMR Gene Expression",  cex.axis =1.9, cex.lab=1.9, cex.main=1.5)
legend("topleft", legend=c("WT","Mutation"),col=c("dodgerblue3","firebrick3"),pch=19, bty="n",y.intersp=0.3,x.intersp=0.3,text.width=0.1, cex=1.2)
#labelit("A",cex=1.5, font=2)

cor.test(BR,MR)

# Panel B: RRG and BER pathway scores based on TP53
plot(BR,RG,pch=19,col=col3, type = "p", xlab="BER Gene Expression",ylab="RRG Gene Expression", cex.axis =1.9, cex.lab=1.9, cex.main=1.5)
legend("topleft", legend=c("WT","Mutation"),col=c("dodgerblue3","firebrick3"),pch=19, bty="n",y.intersp=0.3,x.intersp=0.3,text.width=0.1, cex=1.2)
#labelit("B",cex=1.5, font=2)

cor.test(BR,RG)

# Panel C: RRG and MMR pathway scores based on TP53
plot(MR,RG,pch=19,col=col3, type = "p", xlab="MMR Gene Expression",ylab="RRG Gene Expression",cex.axis =1.9, cex.lab=1.9, cex.main=1.5)
legend("topleft", legend=c("WT","Mutation"),col=c("dodgerblue3","firebrick3"),pch=19, bty="n",y.intersp=0.3,x.intersp=0.3,text.width=0.1, cex=1.2)
#labelit("C",cex=1.5, font=2)
cor.test(MR,RG)

#strified the Based like based on TP53
META1 <- META[META$PAM %in% c("Basal-like"),]

BRR<-apply(BER[set2,],2,mean,na.rm=TRUE)[rownames(META1)]
MRR<-apply(MMR[setx,],2,mean,na.rm=TRUE)[rownames(META1)]
RGG<-apply(RRG[sety,],2,mean,na.rm=TRUE)[rownames(META1)]

META1$TP53 <- factor(META1$TP53, levels = c("Wild Type", "Mutation"))
col5 = ifelse(META1$TP53=="Mutation","firebrick3","dodgerblue3")


# Panel D: MMR and BER of TNBC based on TP53
plot(BRR,MRR,pch=19,col=col5, type = "p", xlab="BER Gene Expression",ylab="MMR Gene Expression",cex.axis =1.9, cex.lab=1.8, cex.main=1.8, xlim=c(-1.5, 2.5), ylim=c(-1.5, 2.5))
legend("topleft", legend=c("WT","Mutation"),col=c("dodgerblue3","firebrick3"),pch=19, bty="n",y.intersp=0.3,x.intersp=0.3,text.width=0.1, cex=1.2)
#labelit("D",cex=1.5, font=2)
cor.test(BRR,MRR)



# Panel E: RRG and BER of TNBC based on TP53
plot(BRR,RGG,pch=19,col=col5, type = "p", xlab="BER Gene Expression",ylab="RRG Gene Expression",cex.axis =1.9, cex.lab=1.8, cex.main=1.5, xlim=c(-1.5, 2.5), ylim=c(-1.5, 2.5)) 
legend("topleft", legend=c("WT","Mutation"),col=c("dodgerblue3","firebrick3"),pch=19, bty="n",y.intersp=0.3,x.intersp=0.3,text.width=0.1, cex=1.2)
#labelit("E",cex=1.5, font=2)


# Panel F: RRG and MMR of TNBC based on TP53
plot(MRR,RGG,pch=19,col=col5, type = "p", xlab="MMR Gene Expression",ylab="RRG Gene Expression", cex.axis =1.9, cex.lab=1.8, cex.main=1.5, xlim=c(-1.5, 2.5), ylim=c(-1.5, 2.5))
legend("topleft", legend=c("WT","Mutation"),col=c("dodgerblue3","firebrick3"),pch=19, bty="n",y.intersp=0.3,x.intersp=0.3,text.width=0.1, cex=1.2)
#labelit("F",cex=1.5, font=2)





```



```{r}



layout(matrix(c(1,4,2,5,3,6),2,3))
par(mar=c(5,6,2,1))
# Panel B:boxplot of BER based on TP53 Status
boxplot(BR~META$TP53,col=col3,xlab="P53 Status", ylab="BER Gene Expression", cex.axis =1.9, cex.lab=2, cex.main=1.8, ylim =c(-2,4.5))
#labelit("A",cex=2, font=2)
# ttest Statstical  
t9 <- t.test(BR~I(META$TP53=="Wild Type")) #p<0.001
t9
# for pvals
axis(3,at=1:2,line=-2.5,lab=c("",""),tcl=2)
text(x=1.5,y=444,label="p=2.2e-16",srt=4,adj=30,cex=3)

# Panel D:boxplot of MMR based on TP53 Status
boxplot(MR~META$TP53,col=col3,xlab="P53 Status", ylab="MMR Gene Expression", cex.axis =1.9, cex.lab=2, cex.main=1.8, ylim=c(-2,4.5))
#labelit("B",cex=2, font=2)
t4 <- t.test(MR~I(META$TP53=="Wild Type")) #p<0.001
t4
# for pvals
axis(3,at=1:2,line=-2.5,lab=c("",""),tcl=2)
text(x=1.5,y=444,label="p=2.2e-16",srt=4,adj=30,cex=3)

# Panel F:boxplot of RGG based on TP53 Status
boxplot(RG~META$TP53,col=col3,xlab="P53 Status", ylab="RRG Gene Expression", cex.axis =1.9, cex.lab=2, cex.main=1.8, ylim=c(-2,4.5))
#labelit("C",cex=2, font=2)
t5 <- t.test(RG~I(META$TP53=="Wild Type")) #p<0.001
t5
# for pvals
axis(3,at=1:2,line=-2.5,lab=c("",""),tcl=2)
text(x=1.5,y=444,label="p=2.2e-16",srt=4,adj=30,cex=3)


# Panel B:boxplot of BER of TNBC based on TP53 Status
boxplot(BRR~META1$TP53,col=col5,xlab="P53 Status", ylab="BER Gene Expression",cex.axis =1.9, cex.lab=2, cex.main=1.8, ylim =c(-2,4.5))
#labelit("D",cex=2, font=2)
t6 <- t.test(BRR~I(META1$TP53=="Wild Type")) #p<0.001
t6
# for pvals
axis(3,at=1:2,line=-2.5,lab=c("",""),tcl=2)
text(x=1.5,y=333,label="p=2.2e-16",srt=4,adj=30,cex=3)


# Panel D:boxplot of MMR of TNBC based on TP53 Status
boxplot(MRR~META1$TP53,col=col5,xlab="P53 Status", ylab="MMR Gene Expression",cex.axis =1.9, cex.lab=2, cex.main=1.8, ylim=c(-2,4.5))
#labelit("E",cex=2, font=2)
t7 <- t.test(MRR~I(META1$TP53=="Wild Type")) #p<0.001
t7
# for pvals
axis(3,at=1:2,line=-2.5,lab=c("",""),tcl=2)
text(x=1.5,y=444,label="p=2.2e-16",srt=4,adj=30,cex=3)


# Panel H3:boxplot of RGG of TNBC based on TP53 Status
boxplot(RGG~META1$TP53,col=col5,xlab="P53 Status", ylab="RRG Gene Expression",cex.axis =1.9, cex.lab=2, cex.main=1.8, ylim=c(-2,4.5))
#labelit("F",cex=2, font=2)
t8 <- t.test(RGG~I(META1$TP53=="Wild Type")) #p<0.001
t8
# for pvals
axis(3,at=1:2,line=-2.5,lab=c("",""),tcl=2)
text(x=1.5,y=444,label="p=2.2e-16",srt=4,adj=30,cex=3)

```

```{r}
## add mut heatmap
xx<-MUTS[,c("TP53","MDM4",
        #"MCM6","MCM2",
        "UNG","TDG","POLB","NEIL3","LIG1","PARP1","APEX1","MUTYH",
        "MSH2","MSH3","MSH6","MLH3","PMS2")]
xx<-as.matrix(xx)
xx[xx=="MUT;"]<-"Mutation"
xx[xx=="HOMDEL;"]<-"Deep Deletion"
xx[xx=="MUT; HOMDEL;"]<-"Mutation/Deep Deletion"
xx[xx=="HOMDEL; MUT;"]<-"Mutation/Deep Deletion"
xx[xx=="AMP;"]<-"Amplification"
xx[xx==""] <- "Wild Type"


setB<-c("PARP1","MBD4","POLD1","LIG1","PCNA","NEIL3",
        "FN1","UNG","TDG","POLD3","MUTYH")
setM<-c("EXO1","RFC2","RFC4","MSH6","MSH2","SSBP1","POLD2",
        "PMS1","PMS2","LIG4","RFC3","RPA1","RPA2")
setR<-c("BUB1","BUB3","CDC25B","CDC25A","BIRC5","CCNE2","CCNB1","CCNB2","CCNF","CDC20","CDC6","CDCA3","CENPE","CENPF","CKS1B","DTL","MCM2","NCN6","PLK1")

setB<-intersect(setB,rownames(BER))
setM<-intersect(setM,rownames(MMR))
setR<-intersect(setR,rownames(RRG))
x1<-t(BER[setB,])
x2<-t(MMR[setM,])
x3<-t(RRG[setR,])

x1<-na.omit(x1)
rw<-rownames(x1)
ra<-META[rw,"PAM"]
rw<-rw[!is.na(ra)]

x1<-x1[rw,]
x2<-x2[rw,]
x3<-x3[rw,]
xx<-xx[rw,]

hc<-hclust(dist(x1)+dist(x2)+dist(x3)+3*dist(xx!="Wild Type"))
col<-colorRamp2(c(-3,0,3),c("blue","white","red"))

h1<-Heatmap(x1,show_row_names = FALSE,
            column_title="Base Excision Repair",
            name="mRNA\nZ-score",cluster_rows=hc,col=col,
            #top_annotation = ta,
            column_title_gp=gpar(fontsize = 16))

ra<-META[rw,"PAM"]
lv<-1:4;names(lv)<-levels(META$PAM); lv<-list(PAM= c("Basal-like"= "dodgerblue3","HER2"= "firebrick3", "Luminal A"="tan1","Luminal B"="limegreen"))
ra<-rowAnnotation("PAM"=ra,col=lv)                 

h2<-Heatmap(x2,show_row_names = FALSE,
            column_title="Mismatch Repair",
            cluster_rows=hc,col=col,
            show_heatmap_legend=TRUE,
            column_title_gp=gpar(fontsize = 16))

h3<-Heatmap(x3,show_row_names = FALSE,
            column_title="Replication Related Genes",
            cluster_rows=hc,col=col,
            show_heatmap_legend=TRUE,
            column_title_gp=gpar(fontsize = 16))

mcol<-c("Wild Type" = "white",
        "Mutation" = "lightblue",
        "Deep Deletion" = "blue",
        "Mutation/Deep Deletion" = "navyblue",
        "Amplification"="red")
ta<-rep(c("p53","DNA Licensing","BER","DDR"),times=c(1,2,8,5))
ta<-columnAnnotation(Gene=ta,
       col=list("Gene"=c("p53"="tomato",
                         "DNA Licensing"="dodgerblue",
                         "BER"="forestgreen",
                         "DDR"="purple")),
       show_annotation_name=FALSE)
h4<-Heatmap(xx[,-1],show_row_names = FALSE,col=mcol,
            name="Mutations", 
            cluster_columns=FALSE,#top_annotation = ta,
            column_title="DNA Repair Alterations",
            column_title_gp=gpar(fontsize = 16))


h5<-Heatmap(xx[,1],show_row_names = FALSE,col=mcol,
            name="TP53",show_heatmap_legend=FALSE)

rp<-HeatmapAnnotation(TMB=anno_barplot(META[rw,"MUT"]),
                      which="row")

#layout(cbind(1,2,3,4:6))
#par(cex.lab=2)
par(mar=c(2,7,2,2))
#plot.new();labelit("E",cex=2,xpd=NA,font=2)

draw(h1+h2+h5+ra+rp+h4,
     newpage=FALSE,
     merge_legend=TRUE)

decorate_heatmap_body("mRNA\nZ-score", {
   grid.lines(c(-.125,-.125,4.10,4.10,-.1), 
              c(0.33, 0,0,.33,.33), 
               gp = gpar(lty = 2, lwd = 4,fill=NA,col="yellow"))})



```


```{r}
# RRG heatmap with PAM50
h3<-Heatmap(x3,show_row_names = FALSE,
            column_title="Replication Related Genes",
            cluster_rows=hc,col=col,
            show_heatmap_legend=TRUE,
            column_title_gp=gpar(fontsize = 16))

h5<-Heatmap(xx[,1],show_row_names = FALSE,col=mcol,
            name="TP53",show_heatmap_legend=FALSE)
draw(h3+h5+ra+rp,
     newpage=FALSE,
     merge_legend=TRUE)
```


## Session Information

```{r}
sessionInfo()
```
